Author: Olivia Holt

from MEDS course: eds223 week 9 lab

# Overview

Monitoring the distribution and change in land cover types can help us understand the impacts of phenomena like climate change, natural disasters, deforestation, and urbanization. Determining land cover types over large areas is a major application of remote sensing because we are able to distinguish different materials based on their spectral reflectance.

Classifying remotely sensed imagery into landcover classes enables us to understand the distribution and change in landcover types over large areas. There are many approaches for performing landcover classification -- supervised approaches use training data labeled by the user, whereas unsupervised approaches use algorithms to create groups which are identified by the user afterward.

credit: this lab is based on a materials developed by Chris Kibler.

# Task

To use a form of supervised classification, a decision tree classifier. [Decision trees](https://medium.com/@ml.at.berkeley/machine-learning-crash-course-part-5-decision-trees-and-ensemble-models-dcc5a36af8cd) classify pixels using a series of conditions based on values in spectral bands. These conditions (or decisions) are developed based on training data. To create a land cover classification for southern Santa Barbara County based on multi-spectral imagery and data on the location of 4 land cover types:

-   green vegetation\
-   dry grass or soil\
-   urban\
-   water\

## Summary

-   load and process Landsat scene\
-   crop and mask Landsat data to study area\
-   extract spectral data at training sites\
-   train and apply decision tree classifier\
-   plot results

## Data

**Landsat 5 Thematic Mapper**\

-   [Landsat 5](https://www.usgs.gov/landsat-missions/landsat-5)
-   1 scene from September 25, 2007\
-   bands: 1, 2, 3, 4, 5, 7
-   Collection 2 surface reflectance product\

**Study area and training data**

-   polygon representing southern Santa Barbara county
-   polygons representing training sites\
    -   type: character string with land cover type\

# Workflow

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Process data

#### Load packages and set working directory

Will need both `sf` and `terra` to work with vector and raster data. The `rpart` and `rpart.plot` packages will be used to train the classification algorithm and plot the results.\

```{r include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(terra)
library(here)
library(dplyr)
library(rpart)
library(rpart.plot)
library(tmap)

rm(list = ls())
```

#### Load Landsat data

Create a raster stack based on the 6 bands to work with. Each file name ends with the band number (e.g. `B1.tif`).

```{r include=TRUE}
# list files for each band, including the full file path
filelist <- list.files("./data/landsat-data/", full.names = TRUE)

filelist
# read in and store as a raster stack
landsat <- rast(filelist)

landsat
# update layer names to match band

names(landsat) <- c('blue','green','red','NIR','SWIR1','SWIR2')

landsat

# plot true color image

plotRGB(landsat, r = 3, g = 2, b = 1, stretch = 'lin')
```

#### Load study area

Constrain the analysis to the southern portion of the county to match the location of training data.

```{r include=TRUE}

# read in shapefile for southern portion of SB county
SB_county_south <- st_read("./data/studyarea_training-data/SB_county_south.shp")

# project to match the Landsat data
SB_county_south <- st_transform(SB_county_south, crs = st_crs(landsat))

plot(SB_county_south)
```

#### Crop and mask Landsat data to study area

Now, we can crop and mask the Landsat data to our study area. This reduces the amount of data we'll be working with and therefore saves computational time. We can also remove any objects we're no longer working with to save space.

```{r include=TRUE}

# crop Landsat scene to the extent of the SB county shapefile
landsat_cropped <- crop(landsat, SB_county_south)

# mask the raster to southern portion of SB county
landsat_masked <- mask(landsat_cropped, SB_county_south)

# remove unnecessary object from environment
rm(landsat, landsat_cropped, SB_county_south) #keeping only the masked data
```

#### Convert Landsat values to reflectance

Now we need to convert the values in our raster stack to correspond to reflectance values. To do so, we need to remove erroneous values and apply any [scaling factors](https://www.usgs.gov/faqs/how-do-i-use-scale-factor-landsat-level-2-science-products#:~:text=Landsat%20Collection%202%20surface%20temperature,the%20scale%20factor%20is%20applied.) to convert to reflectance.\

In this case, we are working with [Landsat Collection 2](https://www.usgs.gov/landsat-missions/landsat-collection-2). The valid range of pixel values for this collection 7,273-43,636, with a multiplicative scale factor of 0.0000275 and an additive scale factor of -0.2. So we reclassify any erroneous values as `NA` and update the values for each pixel based on the scaling factors. Now the pixel values should range from 0-100%.

```{r include=TRUE}

summary(landsat_masked)
# reclassify erroneous values as NA
rcl <- matrix(c(-Inf, 7273, NA,
         43636, Inf, NA),
       ncol = 3, byrow = TRUE)

landsat <- classify(landsat_masked, rcl = rcl)

# adjust values based on scaling factor
landsat <- (landsat * 0.0000275 - 0.2)*100

# plot true color image to check results

plotRGB(landsat, r = 3, g = 2, b = 1, stretch = 'lin')

# check values are 0 - 100

summary(landsat)

```

## Classify image

#### Extract reflectance values for training data

We will load the shapefile identifying different locations within our study area as containing one of our 4 land cover types. We can then extract the spectral values at each site to create a data frame that relates land cover types to their spectral reflectance.

```{r include=TRUE}
# read in and transform training data
training_data <- st_read("./data/studyarea_training-data/trainingdata.shp") %>% 
  st_transform(., crs = st_crs(landsat)) #'.,' the dot is there in place of the data input, stylistic placeholder. not needed

# extract reflectance values at training sites

training_data_values <- extract(landsat, training_data, df = TRUE)

# convert training data to data frame

training_data_attributes <- training_data %>%
  st_drop_geometry()

# join training data attributes and extracted reflectance values

SB_training_data <- left_join(training_data_values, training_data_attributes, by = c("ID" = "id")) %>%
  mutate(type = as.factor(type))# converting landcover type to factor

```

#### Train decision tree classifier

To train our decision tree, we first need to establish our model formula (i.e. what our response and predictor variables are). The `rpart` function implements the [CART algorithm](https://medium.com/geekculture/decision-trees-with-cart-algorithm-7e179acee8ff). The `rpart` function needs to know the model formula and training data you would like to use. Because we are performing a classification, we set `method = "class"`. We also set `na.action = na.omit` to remove any pixels with `NA`s from the analysis.\

To understand how our decision tree will classify pixels, we can plot the results. The decision tree is comprised of a hierarchy of binary decisions. Each decision rule has 2 outcomes based on a conditional statement pertaining to values in each spectral band.

```{r include=TRUE}
# establish model formula

SB_formula <- type ~ red + green + blue + NIR + SWIR1 + SWIR2

# train decision tree

SB_decision_tree <- rpart(formula = SB_formula,
                          data = SB_training_data,
                          method = "class",
                          na.action = na.omit)

# plot decision tree

prp(SB_decision_tree)
```

#### Apply decision tree

Now that we have created our decision tree, we can apply it to our entire image. The `terra` package includes a `predict()` function that allows us to apply a model to our data. In order for this to work properly, the names of the layers need to match the column names of the predictors we used to train our decision tree. The `predict()` function will return a raster layer with integer values. These integer values correspond to the *factor levels* in the training data. To figure out what category each integer corresponds to, we can inspect the levels of our training data.

```{r include=TRUE}
# classify image based on decision tree

SB_classification <- predict(landsat, SB_decision_tree, type = "class", na.rm = TRUE)

# inspect level to understand the order of classes in prediction

levels(SB_training_data$type)

```

#### Plot results

Now we can plot the results and check out our land cover map!

```{r}
# plot results

tm_shape(SB_classification) +
  tm_raster(col.scale = tm_scale_categorical(values = c("#8DB580", "#F2DDA4", "#7E8987", "#6A8EAE")),
            col.legend = tm_legend(labels = c("green vegetation", "soil/dead grass", "urban", "water"),
                                   title = "Landcover type")) +
  tm_layout(legend.position = c("left", "bottom"))

```
